Iterative construction of replicated designs based on Sobol' sequences

### 2016-06-29 crass/crassnew.tex

The replication procedure described in \cite{Mara,Tissot} allows the estimation of all first-order Sobol' indices with only two replicated designs of order $1$. This procedure has the major advantage of reducing the number of model evaluations, evaluating only on designs $\mathcal{P}$ and $\mathcal{P}'$ regardless of the input space dimension. However, Sobol' indices estimates may still not be accurate enough if designs $\mathcal{P}$ and $\mathcal{P}'$ do not explore the input space properly.

\bigskip

@@

\begin{e-definition}%[$(t,m,s)-net$]
Let $\mathcal{A}$ be the set of all subintervals $A\subset\cube$ where $A=\prod_{j=1}^s [\alpha_jb^{-\gamma_j},(\alpha_j+1)b^{-\gamma_j})$, with integers $s\geq 1$, $b\geq 2$, $\gamma_j\geq 0$, and $b^{\gamma_j}>\alpha_j\geq 0$. For $m\geq t\geq 0$, the point set $\mathcal{P}\in\cube$ with $b^m$ points is a $(t,m,s)-net$ in base $b$ if every $A$ with volume $b^{t-m}$ contains $b^t$ points of $\mathcal{P}$.
\end{e-definition}

@@

With the multiplicative approach, the size of designs $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ is multiplied by $2$ each time $\ell$ is increased by one. This growth rate may be inadequate for some applications. The additive approach presented in this section is attractive due to a slower size growth. Given an initial choice of $r\geq 1$ that specifies the size of $B_0$ and ${B'}_0$, only $|B_0|=2^r$ points are added to both designs at each step. The main drawback of this approach is that $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ do not inherit the structure of a Sobol' sequence when $\ell\geq 1$. Nevertheless, both designs display good space-filling properties as shown in the next section.

\bigskip

@@

At step $\ell=0$, $B_0=\{\vx_i^{(0)}\}_{i=0}^{2^r-1}$ and ${B'}_0=\{{\vx'}_i^{(0)}\}_{i=0}^{2^r-1}$ are generated as follows:

for each $i=0,\dots,2^r-1$, point $\vx_i^{(0)} = (x_{i,1}^{(0)},\dots,x_{i,s}^{(0)})^\intercal$ and ${\vx'}_i^{(0)} = ({x'}_{i,1}^{(0)},\dots,{x'}_{i,s}^{(0)})^\intercal$ are obtained by linearly transforming $\vx_i$,
\begin{equation}
\label{linear.scrambl}
\begin{aligned}
(x_{i,j,1}^{(0)},\dots,x_{i,j,r}^{(0)})^\intercal &= L (x_{i,j,1},\dots,x_{i,j,r})^\intercal,\qquad j= 1,\dots,s\, , \\
({x'}_{i,j,1}^{(0)},\dots,{x'}_{i,j,r}^{(0)})^\intercal &= L' (x_{i,j,1},\dots,x_{i,j,r})^\intercal,\qquad j= 1,\dots,s\, , 
\end{aligned}
\end{equation}
where $ x_{i,j}^{(0)}=\sum_{k = 1}^r x_{i,j,k}^{(0)}2^{-k}$, $ {x'}_{i,j}^{(0)}=\sum_{k = 1}^r {x'}_{i,j,k}^{(0)}2^{-k}$ and $L$, $L'$ are two distinct full rank lower triangular matrices of size $r \times r$ over $\mathbb{F}_2$. Left multiplications by matrices $L$ and $L'$ in \eqref{linear.scrambl} are called linear scramblings \cite{Matousek}.

At step $\ell \geq 1$, using definition \eqref{group.operation} we construct:
\begin{equation}
B_\ell=\{\vx_i^{(0)}\oplus\ve_\ell\}_{i=0}^{2^r-1}, \qquad {B'}_\ell=\{{\vx'}_i^{(0)}\oplus{\ve'}_\ell\}_{i=0}^{2^r-1}.
\label{dig.shift}
\end{equation}
where $\ve_\ell \in \left(\mathbb{F}_2^{r}\right)^{s} \setminus \mathcal{P}_{\ell-1}$ and ${\ve'}_\ell \in \left(\mathbb{F}_2^{r}\right)^{s} \setminus \mathcal{P'}_{\ell-1}$. Indeed, $B_0$ is a subgroup of $\left(\mathbb{F}_2^{r}\right)^{s}$. Therefore, we construct $B_\ell=B_0 \oplus \ve_\ell$ as a coset of $B_0$ under the right choice of $\ve_\ell$. The same reasoning applies to ${B'}_0$.  Additions by vectors $\ve_j$ and ${\ve'}_j$ in \eqref{dig.shift} are referred as digital shifts.

There are exactly $2^{rs}$ different choices of $\ve_\ell$ and ${\ve'}_\ell$ in $\left(\mathbb{F}_2^{r}\right)^{s}$, and $2^{r(r-1)/2}$ choices of $L$ and $L'$. The maximum number of iterations $\ell$ that can be performed equals the number of cosets, $2^{r(s-1)}$. If this maximum is reached, designs $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ both correspond to a full grid design of $2^{rs}$ points.

@@

\begin{lemma}
Under the additive construction $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ are two replicated designs of order $1$, $\ell \geq 0$.
\end{lemma}
\textit{Proof}:For each $\ell\geq 1$, digital shift and scrambling operations carried out in equations (\ref{linear.scrambl}) and (\ref{dig.shift}) are bijections from $\mathbb{F}_2^r$ and $\left(\mathbb{F}_2^{r}\right)^{s}$. Hence, same coordinates of $B_\ell$ and ${B'}_\ell$ will contain the same set of $2^r$ values.

@@

We name by design $M$ (resp. $A$) one of the two replicated designs constructed with the multiplicative (resp. additive) approach, and set the input space dimension $s$ to $6$. For design $M$, we study the properties of $\mathcal{P}_8,\dots,\mathcal{P}_{12}$, and for design $A$, we set the initial value to $r=8$ and study the properties of $\mathcal{P}_0,\dots,\mathcal{P}_{15}$. For both $M$ and $A$ we generate $100$ random designs whose average criteria are presented in Figure \ref{results.crit}. While each of the $100$ designs $M$ is independently randomized applying Owen's scrambling \cite{OwenScrambling}, designs $A$ are constructed by randomly selecting matrix $L$ (uniformly among all $2^{r(r-1)/2}$ choices) and vectors $\ve_\ell$ used in equation (\ref{dig.shift}). In addition, we also show in Figure \ref{results.crit} the results obtained with a Latin Hypercube design (LH design) with an equal number of points as in design $A$. The x-axis corresponds to the number of design points $N$ and the y-axis to the value of the criterion.

As expected, design $M$ performs the best and the estimated slope of the discrepancy curve ($-0.72$) falls within the expected range for Sobol' sequences \cite{Morokoff}. Although design $A$ does not perform as well as design $M$, it outperforms a classic Latin Hypercube design for both criteria while allowing $15$ refinement steps from $2^8$ to $2^{12}$ points, instead of only $4$ steps for design $M$.
\begin{figure}
\caption{Log-log graph of averaged $L_2$ star discrepancy (left) and maximin (right) for designs $M$ and $A$ over $100$ repetitions.}
\label{results.crit}
\centering
\includegraphics[scale=0.4]{discrep.pdf}
\hspace*{0.1cm}
\includegraphics[scale=0.4]{mindist.pdf}
\end{figure}

>>>>

The replication procedure described in \cite{Mara,Tissot} allows the estimation of all first-order Sobol' indices with only two replicated designs of order $1$. This procedure has the major advantage of reducing the number of model evaluations, evaluating only on designs $\mathcal{P}$ and $\mathcal{P}'$ regardless of the input space dimension. However, Sobol' indices estimates may still not be accurate enough if designs $\mathcal{P}$ and $\mathcal{P}'$ do not explore the input space properly.
\bigskip

@@

\begin{e-definition}%[$(t,m,s)-net$]
Let $\mathcal{A}$ be the set of all elementary intervals $A\subset \cube$ where $A=\prod_{j=1}^s [\alpha_jb^{-\gamma_j},(\alpha_j+1)b^{-\gamma_j})$, with integers $s\geq 1$, $b\geq 2$, $\gamma_j\geq 0$, and $b^{\gamma_j}>\alpha_j\geq 0$. For $m\geq t\geq 0$, the point set $\mathcal{P}\in\cube$ with $b^m$ points is a $(t,m,s)-net$ in base $b$ if every $A$ with volume $b^{t-m}$ contains $b^t$ points of $\mathcal{P}$.
\end{e-definition}

@@

With the multiplicative approach, the size of designs $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ is multiplied by $2$ each time $\ell$ is increased by one. This growth rate may be inadequate for some applications. The additive approach presented in this section is attractive due to a slower size growth. Given an initial choice of $r\geq 1$ that specifies the size of $B_0$ and ${B'}_0$, only $|B_0|=2^r$ points are added to both designs at each step. The main drawback of this approach is that $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ do not inherit the structure of a Sobol' sequence when $\ell\geq 1$. Nevertheless, both designs display good space-filling properties as shown in the next section.
\bigskip

@@

At step $\ell=0$, $B_0=\{\vx_i^{(0)}\}_{i=0}^{2^r-1}$ and ${B'}_0=\{{\vx'}_i^{(0)}\}_{i=0}^{2^r-1}$ are generated as follows: for each $i=0,\dots,2^r-1$, points $\vx_i^{(0)} = (x_{i,1}^{(0)},\dots,x_{i,s}^{(0)})^\intercal$ and ${\vx'}_i^{(0)} = ({x'}_{i,1}^{(0)},\dots,{x'}_{i,s}^{(0)})^\intercal$ are obtained by linearly transforming $\vx_i$,
\begin{equation}
\label{linear.scrambl}
\begin{aligned}
(x_{i,j,1}^{(0)},\dots,x_{i,j,r}^{(0)})^\intercal &= L (x_{i,j,1},\dots,x_{i,j,r})^\intercal,\qquad j= 1,\dots,s\, , \\
({x'}_{i,j,1}^{(0)},\dots,{x'}_{i,j,r}^{(0)})^\intercal &= L' (x_{i,j,1},\dots,x_{i,j,r})^\intercal,\qquad j= 1,\dots,s\, , 
\end{aligned}
\end{equation}
where $ x_{i,j}^{(0)}=\sum_{k = 1}^r x_{i,j,k}^{(0)}2^{-k}$, $ {x'}_{i,j}^{(0)}=\sum_{k = 1}^r {x'}_{i,j,k}^{(0)}2^{-k}$ and $L$, $L'$ are two distinct full rank lower triangular matrices of size $r \times r$ over $\mathbb{F}_2$. Left multiplications by $L$ and $L'$ in \eqref{linear.scrambl} are called linear scramblings \cite{Matousek}.

At step $\ell \geq 1$, using addition as defined in \eqref{group.operation} we construct:
\begin{equation}
B_\ell=\{\vx_i^{(0)}\oplus\ve_\ell\}_{i=0}^{2^r-1}, \qquad {B'}_\ell=\{{\vx'}_i^{(0)}\oplus{\ve'}_\ell\}_{i=0}^{2^r-1}.
\label{dig.shift}
\end{equation}
where $\ve_\ell \in \left(\mathbb{F}_2^{r}\right)^{s} \setminus \mathcal{P}_{\ell-1}$ and ${\ve'}_\ell \in \left(\mathbb{F}_2^{r}\right)^{s} \setminus \mathcal{P'}_{\ell-1}$. Indeed, $B_0$ is a subgroup of $\left(\mathbb{F}_2^{r}\right)^{s}$. Therefore, we construct $B_\ell=B_0 \oplus \ve_\ell$ as a coset of $B_0$ under the right choice of $\ve_\ell$. The same reasoning applies to ${B'}_0$.  Additions by vectors $\ve_j$ and ${\ve'}_j$ in \eqref{dig.shift} are referred as digital shifts.

There are exactly $2^{rs}$ different choices of $\ve_\ell$ and ${\ve'}_\ell$ in $\left(\mathbb{F}_2^{r}\right)^{s}$, and $2^{r(r-1)/2}$ choices of $L$ and $L'$. The maximum number of iterations $\ell$ that can be performed equals the number of cosets, $2^{r(s-1)}$. If this maximum is reached, designs $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ both correspond to a full grid design of $2^{rs}$ points with $2^r$ distinct values per input.

@@

\begin{lemma}
Under the additive construction $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ are two replicated designs of order $1$, $\ell \geq 0$.
\end{lemma}
\textit{Proof}: For each $\ell\geq 1$, digital shift and scrambling operations carried out in equations (\ref{linear.scrambl}) and (\ref{dig.shift}) are bijections from $\mathbb{F}_2^r$ and $\left(\mathbb{F}_2^{r}\right)^{s}$. Hence, the same coordinates of $B_\ell$ and ${B'}_\ell$ will contain the same set of $2^r$ values.

@@

We name by design $M$ (resp. $A$) one of the two replicated designs constructed with the multiplicative (resp. additive) approach, and set the input space dimension $s$ to $6$. For design $M$, we study the properties of $\mathcal{P}_8,\dots,\mathcal{P}_{12}$, and for design $A$, we set the initial value to $r=8$ and study the properties of $\mathcal{P}_0,\dots,\mathcal{P}_{15}$. For both $M$ and $A$ we generate $100$ designs whose average criteria are presented in Figure \ref{results.crit}. While each of the $100$ designs $M$ is independently randomized applying Owen's scrambling \cite{OwenScrambling}, designs $A$ are constructed by randomly selecting matrix $L$ (uniformly among all $2^{r(r-1)/2}$ choices) and vectors $\ve_\ell$ used in equation (\ref{dig.shift}). In addition, we also show in Figure \ref{results.crit} the results obtained with a Latin Hypercube design (LH design) with an equal number of points as in design $A$. LH designs are the standard designs used in the replication procedure to estimate first-order indices. The x-axis corresponds to the number of design points $N$ and the y-axis to the value of the criterion.

As expected, design $M$ performs the best and the estimated slope of the discrepancy curve ($-0.72$) falls within the expected range for Sobol' sequences \cite{Morokoff}. Although design $A$ does not perform as well as design $M$, it outperforms a classic LH design for both criteria while allowing $15$ refinement steps from $2^8$ to $2^{12}$ points, instead of only $4$ steps for design $M$.
\begin{figure}
\caption{Log-log graph of averaged $L_2$ star discrepancy (left) and maximin (right) for designs $M$ and $A$ over $100$ repetitions.}
\label{results.crit}
\centering
\includegraphics[scale=0.38]{discrep.pdf}
\hspace*{0.1cm}
\includegraphics[scale=0.38]{mindist.pdf}
\end{figure}
